The First Level of Super Mario Bros.~is Easy with Lexicographic Orderings and Time Travel {\normalsize \ldots after that it gets a little tricky.}

### 2013-03-17 tasbot/paper/paper.tex
Section completed in paper

I literally spent about six weekends on this problem. First, the
overhang is set up so that enemies emerge from it just as you arrive.
This means that loose play (imperfect enemy avoidance) tends to get
you killed right here. Mario has found a number of different solutions
to this problem, from waiting, to kicking turtles from the left to
clear out the enemies, to timing his jump perfectly (it is possible)
to stomp the turtle after bouncing his head off the ceiling.

@@

@@

>>>>

I literally spent about six weekends and thousands of hours of CPU on
this problem. First, the overhang is set up so that enemies emerge
from it just as you arrive. This means that loose play (imperfect
enemy avoidance) tends to get you killed right here. Mario has found a
number of different solutions to this problem, from waiting, to
kicking turtles from the left to clear out the enemies, to timing his
jump perfectly (it is possible!) to stomp the turtle after bouncing his
head off the ceiling. It's fun to see the different approaches and is
a good benchmark for whether the search algorithm is actually doing a
good job at tactics.

@@

Immediately after this is an important test of longer-term planning.
Most of my early solutions would jump up on this ledge with 4 coins,
since the score is a component of many lexicographic
orderings\footnote{Maybe even all of them, since it should be globally
  obeyed; it's therefore a valid extension to any ordering that
  doesn't already include it. It's not necessarily a tight extension,
  however, so it may be excluded for that reason, or because during
  initialization it is not actually zero and so not globally obeyed. I
  never checked this stuff because I wanted to avoid any temptation to
  overfit the learning algorithm to the particulars of any game.} and
coins give you points. Down below is undesirable in the short term
because of the enemies. Then Mario would feel stuck and just hump up
against the wall jumping in this tiny dead end until time ran out. The
game contains some bugs (you have probably seen them) that allow the
screen to be scrolled without Mario actually progressing; these little
mini scrolls, plus the givens like the frame counter and music cursor,
prevent Mario from getting out of the local maximum. This is extremely
frustrating. I decided to add some longer-term planning into the
search procedure in order to try to help in these kinds of situations,
as well as the occasional deaths I would see.

Every {\tt CHECKPOINT\_EVERY} frames we save the state, and then every
{\tt TRY\_BACKTRACK\_EVERY} rounds, we do a backtracking phase if we
can. We just need to find a checkpoint at least {\tt
  MIN\_BACKTRACK\_DISTANCE} frames in the past. Call that point
{\tt start} and the current point {\tt now}. The backtracking
routine looks like this:

\begin{itemize}
  \item Let {\tt improveme} be the inputs between {\tt start} and {\tt now}.
  \item Get {\tt replacements}, a set of input sequences we might use
    instead. These are usually based on {\tt improveme}.
  \item Add {\tt improveme} to the set of replacements.
  \item Truncate the movie to {\tt start}.
  \item Now, do the normal playfun loop as though the (truncated) futures
    are whatever our current futures are, and the set of {\tt next}
    sequences are the {\tt replacements} array.
  \item Whatever the best one among those is, keep it. Don't update motifs
    or futures, however.
\end{itemize}

The idea is that we take a large sequence from our recent past, and the
same futures we're already using, and see if that sequence can be improved,
according to our objective functions, and using our same futures. Since
the existing sequence is always one of those, if it is the best one,
then the backtracking phase is a no-op. If we find something better, we
slot it in and continue. So the only risk here is if our objective functions
aren't good (we take as given that they are) and the only cost is time.

Generating the candidate set of replacements uses a bunch of different
techniques. They are:

\paragraph{Random.} We generate a random sequence of the same length as
the {\tt improveme} sequence.

\paragraph{Opposites.} We dualize (swap left with right, up with down, start 
with select, and B with A) and/or reverse random spans of the {\tt
  improveme} sequence. The idea is to try to introduce some variety in
cases where we're really getting stuck. This was specifically in hopes
that Mario would walk off the coin ledge in {\tt 1-2} and then find that the
course was clear below. I felt okay about this since this seems to be a
generic notion (the buttons do have semantics that are common to most NES
games where left and right are opposites), but that may just have been
rationalization since I was getting desperate. It didn't work; see below.

\paragraph{Ablation.} Randomly blanks out buttons from the input. For
example, if we don't need to jump and jumping is slowing us down or
something, this can remove that and make a small improvement to the
sequence.

\paragraph{Chop.} Removes random spans from the input. This iteratively
removes spans as long as the previous span was an improvement (see
below). This can cause the movie to get shorter by removing parts that
weren't contributing anything.\footnote{However, in games where an
  objective function includes something like a frame counter or the
  music cursor, shorter sequences always score lower than
  otherwise equivalent longer sequences.}

We use the following formula to score a potential improvement:

\begin{center}
\includegraphics[width=0.95 \linewidth]{isimprovement}
\end{center}

The {\tt start} state is as discussed, {\tt old\_end} is the state at
{\tt now}, and {\tt new\_end} is the state after the potential
improvement sequence. $\int_s^o$ is the integral of score changes
along the {\tt improveme} path and $\int_s^n$ along the candidate
improvement. The integral is the weighted sum of the objectives
increased minus the weighted sum of the objectives that decreased, at
each step, all summed up. This works better than just computing the
weighted score from e.g. {\tt start} to {\tt old\_end} when they are
separated by many frames (typical backtrack distances are several
hundred frames). The expression $n-o$\footnote{This is not numerical
  minus but minus according to the set of objective functions. Just
  roll with it.} is just the weighted sum of objectives that increased
minus the weighted sum of objectives that decreased between the old
end state and new end state; we have no option of computing the
integral here because we don't have an input sequence that goes from
the old end to the new end (and there almost certainly isn't one). We
require that three conditions hold:
\begin{enumerate}
\item $\int_s^n \geq \int_s^o$,
\item $\int_s^n > 0$
\item $n - o > 0$
\end{enumerate}

If they all do, then the score for the improvement is 
$\int_s^n - \int_s^o + (n - o)$, which is guaranteed to be positive.

Even if we have many possible replacements, we try only the ones with
the best scores; due to the particulars of the implementation there
is not a simple constant describing this maximum, but it's around 220
with the parameter settings used in this paper.

\parameteralert{
  Again! The humiliating appearance of constants! There are many here,
  having to do with the number of potential replacements we use of
  each type, the maximum number we keep, how often we bactrack, and
  how far. I am not totally satisfied with how backtracking works (see
  below), so I don't want to spend too much energy speculating about
  how to reduce the parameter space here; I'd rather replace the
  method wholesale.}

\paragraph{Takes of woe, tales of joy.} 
Backtracking was initially added in order to fix the mario coin ledge
problem in {\tt 1-2}. It did not help with this. Older versions of the
code would have Mario get past this spot, probably by luck, but in
these he would get stuck in an embarrassing way on a small pit later,
or jump straight into an enemy. Most of the approaches that actually
looked solidly good elsewhere were failing badly here. As I started
writing up this paper, on an airplane, I discovered the bug in the
weighting of objective functions described in
Section~\label{sec:objectiveweighting}. On my trip I had let {\tt
  playfun} run with particularly high constants ({\tt NFUTURES} = 50,
lots of backtracking candidates, etc.) and it had spent thousands of
CPU hours playing to this point, grabbing the coins, and then dying of
timeout, then losing all its lives, starting over, and getting to this
point again! After fixing the bug, I tried again, and it was a huge
breakthrough: Mario jumped up to get all the coins, and then almost
immediately jumped back down to continue the level! On his first try
he beat {\tt 1-2} and then immediately jumped in the pit at the
beginning of {\tt 1-3} just as I was starting to feel suuuuuper smart
(Figure~\ref{fig:mario13}). On a scale of OMFG to WTFLOL I was like
{\em whaaaaaaat}?

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.95 \linewidth]{adventureisland}
\end{center}\vspace{-0.1in}
\caption{After making me feel so successful by finally cleanly
  finishing world {\tt 1-2}, Mario immediately jumps into the first
  trivial pit in {\tt 1-3}. I believe what's going on here is that
  this is actually the best future, because resetting back a single
  screen isn't a large loss, and probably the rest of the futures were
  struggling to make progress. This level is very jumpy and probably
  needs more futures to navigate its obstacles. In any case, Mario,
  Touch\'e.}
\label{fig:mario13}
\end{figure}

Seeing this huge improvement changed my idea about what part of the
code needed work (I now believe that simpler search strategies might
work and better lexicographic order generation and weighting is called
for). But, this was already in the midst of writing up the paper, so
instead I spent the CPU time running the current version on a bunch of
games. Those results are described in Section~\ref{sec:results} and
videos are available on \url{http://tom7.org/mario/}. In the next
section I describe some of the performance improvements I made, and
other modifications to the emulator interface.

\section{Performance} \label{sec:performance}

@@

% better to run on n-1 cpus?
